---
layout: topic
title: Data visualization with ggplot2
subtitle: Visualizing data in R with the ggplot2 package
minutes: 30
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(fig.keep='last', echo = TRUE, purl = FALSE, 
                      results = 'hide', message = FALSE)
```

> ### Learning Objectives
>
> -	Visualize data using R ggplot2 package
> -	Build step-by-step complex plots with the ggplot2 package

For more details and additional tutorials on using ggplot2, see the [official documentation](http://docs.ggplot2.org/current/).

Load required packages.

```{r}
library(dplyr)
library(ggplot2)
```

Load or reload NHTS data. 

The `read_csv()` function has a handy `na = c(...)` input that allows us to quickly deal with the many, many definitions of missing that NHTS has.

```{r}
nhts_per <- readr::read_csv("data/nhts_per.csv", 
                            na = c("", "NA", -1, -8, -5, -4)) %>%
  select(HOUSEID, HHSIZE, HHVEHCNT, TIMETOWK, YEARMILE, USEPUBTR)
```

## `ggplot2` Basics
We will make two plots using the `ggplot2` package. `ggplot2` is a
plotting package that makes it simple to create complex plots from data in a
dataframe. It uses default settings, which help create publication quality
plots with a minimal amount of settings and tweaking.

`ggplot2` graphics are built step by step by adding elements.

To build a ggplot we need to:

1. Bind the plot to a specific data frame using the `data` argument.
```{r, eval=FALSE}
ggplot(data = nhts_per)
```

2. Define aesthetics (`aes`), which maps variables in the data to axes on the
plot or to plotting size, shape color, etc. 

Think of `aes()` like the `data_frame()` function we used before.
```{r, eval=FALSE}
ggplot(data = nhts_per, aes(x = TIMETOWK, y = YEARMILE))
```

3. Add `geoms`, which define the graphical representation of the data in the
plot (points, lines, bars). To add a geom to the plot use `+` operator. Anything
you put in the `ggplot()` function can be seen by any geom layers that you add.
```{r first-ggplot}
ggplot(data = nhts_per, aes(x = TIMETOWK, y = YEARMILE)) +
  geom_point()
```

4. (optional) Add better labeling.
```{r}
ggplot(data = nhts_per, aes(x = TIMETOWK, y = YEARMILE)) +
  geom_point() +
  labs(title = 'Yearly Mileage vs Commute Time',
       x = 'Time to Work (mins)',
       y = 'Yearly Mileage')
```

5. (optional) Add a theme if you want.
```{r}
ggplot(data = nhts_per, aes(x = TIMETOWK, y = YEARMILE)) +
  geom_point() +
  labs(title = 'Yearly Mileage vs Commute Time',
       x = 'Time to Work (mins)',
       y = 'Yearly Mileage') +
  theme_bw()
```

Sometimes adding color and transparency adds meaning to the plot. In this case, since we limited the number of rows of the NHTS data to 200, there is not much overlap. But with a lot of points, alpha can be very helpful.

```{r adding-colors}
ggplot(data = nhts_per, aes(x = TIMETOWK, y = YEARMILE)) +
  geom_point(alpha = 0.3, color = "blue") +
  labs(title = 'Yearly Mileage vs Commute Time',
       x = 'Time to Work (mins)',
       y = 'Yearly Mileage') +
  theme_bw()
```

## Barplot
Let's try a barplot. 

```{r}
ggplot(data = nhts_per, aes(x = HHVEHCNT)) +
  geom_bar() 
```

But this looks a bit funny right? What if, for example, NHTS coded "8" to actually be any household with 8 or more vehicles. We can instead treat this variable as categories, which in R are called factors.

```{r}
ggplot(data = nhts_per, aes(x = as.factor(HHVEHCNT))) +
  geom_bar() +
  scale_fill_brewer()
```

Then maybe we want to get even fancier and see how household size plays into this distribution.
```{r}
ggplot(data = nhts_per, aes(x = as.factor(HHVEHCNT), 
                            fill = as.factor(HHSIZE))) +
  geom_bar()
```

And maybe we don't like those colors, so we'll change them.
```{r}
ggplot(data = nhts_per, aes(x = as.factor(HHVEHCNT), 
                            fill = as.factor(HHSIZE))) +
  geom_bar() +
  scale_fill_brewer()
```


## Faceting
`ggplot2` has a special technique called *faceting* that allows to split one
plot into multiple plots based on some factor. Let's take the same graph and add
another dimension. Let's look at household size and number of vehicles BY
transit useage.

```{r}
ggplot(data = nhts_per, aes(x = as.factor(HHVEHCNT), 
                            fill = as.factor(HHSIZE))) +
  geom_bar() +
  scale_fill_brewer() +
  facet_grid(. ~ USEPUBTR)
```

> ### Challenge {.challenge}
>
>  Update the labeling so that people can understand the plot.

```{r}
ggplot(data = nhts_per, aes(x = as.factor(HHVEHCNT), 
                            fill = as.factor(HHSIZE))) +
  geom_bar() +
  scale_fill_brewer() +
  facet_grid(. ~ USEPUBTR) +
  labs(title = 'Household Size by Number of Vehicles by Transit Use',
       x = 'Household Size',
       y = 'Count') +
  guides(fill = guide_legend(title = "Num of Vehs"))
```

Enjoy plotting with ggplot2! Use Google search, other online tutorials, and R's
help functionality to teach yourself more.
