---
layout: topic
title: Censusr
subtitle: Grabbing data in R from the Census API
minutes: 10
---

```{r, echo=FALSE, purl=FALSE}
knitr::opts_chunk$set(fig.keep='last', echo = TRUE, purl = FALSE, 
                      results = 'hide')
```

> ### Learning Objectives
>
> - Grab ACS or Decennial Census data from Census API

To install this package, run the command below. For full details on how to get started using `censusr` on your computer, see the package vignette.

```{r, eval = F}
install.packages("censusr")
vignette("censusr")
```

Load required package.

```{r, message = FALSE}
library(censusr)
```

And then let's get to business! The package works by sending a list of requested variables and a list of geographies. The call below requests the number of households owning 0, 1, 2, 3, or 4 or more vehicles in Wake County, North Carolina (geoid = 37183). We specify that we want this table for 2012 5-year summary level.

```{r}
variable_list <- paste0("B08201_", sprintf("%03d", 2:6),  "E")
call_census_api(
  variable_list, names = c(0:4), geoids = "37183",  
  data_source = "acs", year =  2012, period = 5) 
```

We can use the `allgeos` argument to say that we actually want these variables for all census tracts within Wake County.

```{r}
call_census_api(
  variable_list, names = paste0("est_", c(0:4)), 
  geoids = "37183",  allgeos = "tr",
  data_source = "acs", year =  2012, period = 5)
```

If we want the margins of error on this table instead of the estimates, we can change the variable to call the M type instead of the E type.

```{r}
new_var_list <- paste0("B08201_", sprintf("%03d", 2:6),  "M")
call_census_api(
  new_var_list, names = paste0("moe_", c(0:4)), 
  geoids = "37183",  allgeos = "tr",
  data_source = "acs", year =  2012, period = 5)
```
